---
title: "Diagnostic test assessment"
#author: G.Bisaccia
format: 
  dashboard:
    nav-buttons: 
      - twitter
      - icon: github
        href: https://github.com/giandomd
        aria-label: GitHub
    orientation: rows
    scrolling: false
logo: images/logo.png
theme: journal
fig-width: 10
fig-asp: 0.3
editor_options: 
  chunk_output_type: console
server: shiny
---

```{r}
#| context: setup
#| label: load-packages
#| message: false
library(ggplot2)
library(ggthemes)
library(tidyverse)
library(readxl)
library(scales)
library(DT)
library(gt)
library(dplyr)
library(riskyr)
library(extrafont)
pal <- init_pal(riskyr::pal_rgb)
```

#  {.sidebar}

```{r}
textInput('cond', 'Condition',value = 'CAD')
textInput('pop', 'Population',value = 'Stable angina')
textInput('test', 'Index test',value = 'ETT')
textInput('ref', 'Reference standard',value = 'ICA')
numericInput('sample', 'Sample size', 
            min=1,
            value=100, 
            step=1)
br()
numericInput('sens', 'Sensitivity (%)', 
            min=1, max=100,
            value=80, 
            step=0.1)
numericInput('spec', 'Specificity (%)', 
            min=1, max=100,
            value=80, 
            step=0.1)
numericInput('prev', 'Disease prevalence (%)', 
            min=1, max=99.9,
            value=40, 
            step=0.1)
```

::: {.callout-note collapse="true"}
## Author

Developed by G. Bisaccia, Gabriele D'Annunzio University of Chieti-Pescara, Chieti, Italy [ORCID](https://orcid.org/0000-0003-4606-3093)
:::


# Diagnostic accuracy

##

```{r}
library(bslib)
library(bsicons)
value_box(
  title = "Accuracy",
  value = textOutput("ACC"),
  showcase = bs_icon("file-medical")
)
```

##

```{r}
#value_box(title = "Positive likelihood ratio",value = textOutput("PLR"),showcase = bs_icon("plus-lg"))
value_box(title = "Risk after positive test",value = textOutput("ptpos"),showcase = bs_icon("plus-lg"))
```

```{r}
#value_box(title = "Negative likelihood ratio",value = textOutput("NLR"),showcase = bs_icon("dash-lg"))
value_box(title = "Risk after negative test",value = textOutput("ptneg"),showcase = bs_icon("dash-lg"))
```

```{r}
value_box(
  title = "Positive predictive value", 
  value = textOutput("PPV"),
  showcase = bs_icon("plus-circle")
)
```

```{r}
value_box(
  title = "Negative predictive value",
  value = textOutput("NPV"),
  showcase = bs_icon("dash-circle")
)
```

##

```{r}
#| title: Patient flow diagram
plotOutput('p0')
```

```{r}
#| title: Patients according to test results
plotOutput('p4')
```

# Clinical utility

## {height=60%}

```{r}
plotOutput('pfagan')
```


```{r}
plotOutput('dca')
```

## {height=40%}
### {.toolbar}
::: callout-tip
The Fagan nomogram allows to evaluate how a test result affects the probability of a patient having a certain disease. Starting from a given pre-test risk (e.g. disease prevalence among referred patients), a positive (abnormal) test result will (likely) increase the odds, and a negative (normal) test result will decrease the odds, of a patient having the disease. Evaluating such test probabilities against set thresholds for further testing or action (e.g. patients at ≥ 65% risk should *definitely* be tested, or patients at \< 15% risk should definitely avoid testing) allows to evaluate the clinical value of a test. If the test fails to reclassify patients to a higher (or lower) risk category, it is likely to yield very little additional value over the reference standard.

:::
### {.toolbar}
::: callout-tip
Decision curve analysis allows to evaluate the clinical utility of a given test, beyond diagnostic accuracy metrics. It calculates the net benefit of undergoing an index test over a strategy of referring all patients to the reference standard, and over a strategy of no testing. Please provide a clinically meaningful threshold probability range (see [Vickers et al., BMJ 2016](https://doi.org/10.1136/bmj.i6)).

```{r}
sliderInput("range", "Threshold probability range",
min = 0, max = 100, value = c(15,65))
```
:::

# About

This dashboard allows to evaluate the diagnostic yield and clinical utility of a given test against a reference standard. Provided with metrics of sensitivity, specificity, and sample size and disease prevalence, it calculates accuracy, post-test disease risks, and predictive values. It also shows flow diagrams of patients within the study, and dependence of diagnostic estimates on disease prevalence.

Term definitions and formulas used to derive diagnostic metrics are available at [BMJ Best Practice](https://bestpractice.bmj.com/info/toolkit/learn-ebm/diagnostic-test-studies-assessment-and-critical-appraisal/), which also offers practical tips for evaluating the quality of published diagnostic accuracy studies.

Literature on decision curve analysis for assessment of clinical benefit is available at [http://www.decisioncurveanalysis.org](http://www.decisioncurveanalysis.org).

```{r}
#| context: server

output$PLR <- reactive(
  round(ifelse(input$spec==100, no = input$sens/(100-input$spec), yes = 100),digits = 2)
  )
output$NLR <- reactive(
  round(ifelse(input$spec==0, no = (100-input$sens)/(input$spec), yes = 0),digits = 2)
  )
output$PPV <- reactive(paste(
  round((input$sens*input$prev/(input$sens*input$prev+(100-input$spec)*(100-input$prev)))*100, digits = 1),"%",sep = "")
  )
output$NPV <- reactive(paste(
  round((input$spec*(100-input$prev)/((100-input$sens)*input$prev+(input$spec)*(100-input$prev)))*100,digits = 1), "%",sep = "")
  )
output$ACC <- reactive(
  paste(round((input$sens*input$prev+input$spec*(100-input$prev))/100,digits = 1), "%",sep = "")
  )

output$ptneg <- reactive(
  paste(round(
    (input$prev/(100-input$prev))*
      ifelse(input$spec==0, no = (100-input$sens)/(input$spec), yes = 100)/
      (1+(input$prev/(100-input$prev))*
         ifelse(input$spec==0, no = (100-input$sens)/(input$spec), yes = 100))*100, digits = 1)
    ,"%"
        )
  )

output$ptpos <- reactive(
  paste(round(
    (input$prev/(100-input$prev))*
  ifelse(input$spec==100, no = input$sens/(100-input$spec), yes = 100)/
  (1+(input$prev/(100-input$prev))*
     ifelse(input$spec==100, no = input$sens/(100-input$spec), yes = 100))*100,digits = 1)
    ,"%"
  )
)
  

output$TP <- reactive(round((input$sens/100)*(input$prev/100)*input$sample,digits = 0))
output$FN <- reactive(round((input$prev/100)*(input$sample)*(1-input$sens/100),digits = 0))
output$TN <- reactive(round((input$spec/100)*input$sample*(1-input$prev/100),digits = 0))
output$FP <- reactive(round(input$sample*(1-input$prev/100)*(1-input$spec/100),digits = 0))

output$p0 <- renderPlot(riskyr::riskyr(scen_lbl = 
                        paste(input$cond, 
                              "reclassification by", 
                              input$test),
                      popu_lbl = input$pop,
                      cond_true_lbl = paste(input$cond, "present"),
                      cond_false_lbl = paste(input$cond, "absent"),
                      dec_lbl = paste(input$test, "result"),
                      dec_pos_lbl = "predicts presence", 
                      dec_neg_lbl = "predicts absence",
                      hi_lbl = paste(input$cond, 
                                     "properly diagnosed"),
                      mi_lbl = paste(input$cond, "missed"),
                      fa_lbl = paste(input$cond, "falsely diagnosed"),
                      cr_lbl = paste(input$cond, "properly excluded"),
                      prev = input$prev/100,
                      sens = input$sens/100,
                      spec = input$spec/100, 
                      N = input$sample
                      ) %>% plot(type='prism', col_pal=riskyr::pal_rgb))

output$p1 <- renderPlot(riskyr::riskyr(scen_lbl = 
                        paste(input$cond, 
                              "reclassification by", 
                              input$test),
                      popu_lbl = input$pop,
                      cond_true_lbl = paste(input$cond, "present"),
                      cond_false_lbl = paste(input$cond, "absent"),
                      dec_lbl = paste(input$test, "result"),
                      dec_pos_lbl = "predicts presence", 
                      dec_neg_lbl = "predicts absence",
                      hi_lbl = paste(input$cond, 
                                     "properly diagnosed"),
                      mi_lbl = paste(input$cond, "missed"),
                      fa_lbl = paste(input$cond, "falsely diagnosed"),
                      cr_lbl = paste(input$cond, "properly excluded"),
                      prev = input$prev/100,
                      sens = input$sens/100,
                      spec = input$spec/100, 
                      N = input$sample
                      ) %>% plot(type='curve', col_pal=riskyr::pal_rgb))

output$p2 <- renderPlot(riskyr::riskyr(scen_lbl = 
                        paste(input$cond, 
                              "reclassification by", 
                              input$test),
                      popu_lbl = input$pop,
                      cond_true_lbl = paste(input$cond, "present"),
                      cond_false_lbl = paste(input$cond, "absent"),
                      dec_lbl = paste(input$test, "result"),
                      dec_pos_lbl = "predicts presence", 
                      dec_neg_lbl = "predicts absence",
                      hi_lbl = paste(input$cond, 
                                     "properly diagnosed"),
                      mi_lbl = paste(input$cond, "missed"),
                      fa_lbl = paste(input$cond, "falsely diagnosed"),
                      cr_lbl = paste(input$cond, "properly excluded"),
                      prev = input$prev/100,
                      sens = input$sens/100,
                      spec = input$spec/100, 
                      N = input$sample
                      ) %>% plot(type='tab', col_pal=riskyr::pal_rgb))

output$p3 <- renderPlot(riskyr::riskyr(scen_lbl = 
                        paste(input$cond, 
                              "reclassification by", 
                              input$test),
                      popu_lbl = input$pop,
                      cond_true_lbl = paste(input$cond, "present"),
                      cond_false_lbl = paste(input$cond, "absent"),
                      dec_lbl = paste(input$test, "result"),
                      dec_pos_lbl = "predicts presence", 
                      dec_neg_lbl = "predicts absence",
                      hi_lbl = paste(input$cond, 
                                     "properly diagnosed"),
                      mi_lbl = paste(input$cond, "missed"),
                      fa_lbl = paste(input$cond, "falsely diagnosed"),
                      cr_lbl = paste(input$cond, "properly excluded"),
                      prev = input$prev/100,
                      sens = input$sens/100,
                      spec = input$spec/100, 
                      N = input$sample
                      ) %>% plot(type='bar', col_pal=riskyr::pal_rgb))

output$p4 <- renderPlot({
  
  r <- riskyr::riskyr(scen_lbl = 
                        paste(input$cond, 
                              "reclassification by", 
                              input$test),
                      popu_lbl = input$pop,
                      cond_true_lbl = paste(input$cond, "present"),
                      cond_false_lbl = paste(input$cond, "absent"),
                      dec_lbl = paste(input$test, "result"),
                      dec_pos_lbl = "predicts presence", 
                      dec_neg_lbl = "predicts absence",
                      hi_lbl = paste(input$cond, 
                                     "properly diagnosed"),
                      mi_lbl = paste(input$cond, "missed"),
                      fa_lbl = paste(input$cond, "falsely diagnosed"),
                      cr_lbl = paste(input$cond, "properly excluded"),
                      prev = input$prev/100,
                      sens = input$sens/100,
                      spec = input$spec/100, 
                      N = input$sample) %>% plot(type="icon", col_pal=riskyr::pal_rgb)
                      })

output$pfagan <- renderPlot({
  data.frame(
  ecg_pos=c(input$prev, 
            (input$prev/(100-input$prev))*
              100/
              (1+(input$prev/(100-input$prev))*
                 100)*100),
  ecg_neg=c(input$prev, (input$prev/(100-input$prev))*
              ifelse(input$spec==0, no = (100-input$sens)/(input$spec), yes = 100)/
              (1+(input$prev/(100-input$prev))*
                 ifelse(input$spec==0, no = (100-input$sens)/(input$spec), yes = 100))*100),
  time = c(0, 1)
) %>% ggplot()+
  geom_rect(aes(xmin = time[1], xmax = time[2], ymin = input$range[2], ymax = 100), alpha=0.05, fill="darkred")+
  geom_rect(aes(xmin = time[1], xmax = time[2], ymin = 0, ymax = input$range[1]), alpha=0.05)+
  geom_line(aes(x = time, y = ecg_pos), color="#f1184c", cex=1.1)+
  geom_line(aes(x = time, y = ecg_neg), color="#f1184c", linetype="dashed", cex=1.1)+
  #geom_rect(aes(xmin = time[1], xmax = time[2], ymin = input$range[2], ymax = 100), alpha=0.15, fill="darkred")+
  geom_rect(aes(xmin = time[1], xmax = time[2], ymin = 0, ymax = input$range[1]), alpha=0.15)+
  geom_line(aes(x = time, y = input$range[2]),linetype="dashed", color="black")+
  geom_line(aes(x = time, y = input$range[1]),linetype="dashed", color="black")+
  ylim(0,100)+
  ylab(NULL)+
  xlab("")+
  scale_x_continuous(breaks = NULL)+
  coord_fixed(ratio=0.01)+
  theme_economist_white(gray_bg = F, base_size = 14, base_family = "Atkinson Hyperlegible")+
  ggtitle(label = "Fagan nomogram\n", subtitle  = paste(input$test, " detection of ", input$cond, " in patients with ", input$pop, ", ", input$ref, " as reference", sep = ""))+
  scale_y_continuous(name = "Pre-test probability\n", sec.axis = dup_axis(name = "Post-test probability"))#+xlim(0,1))
})

output$dca <- renderPlot({
  riskyr_result <- riskyr(prev = input$prev/100, sens = input$sens/100, spec = input$spec/100, N = input$sample)
  TP <- riskyr_result$hi
  TN <- riskyr_result$cr
  FP <- riskyr_result$fa
  FN <- riskyr_result$mi
  n <- riskyr_result$N

  seq_range <- seq(input$range[1], input$range[2], by = 0.5)
  n_values <- numeric(length(seq_range))
  
  # Iterate over the chosen interval for p
  for (i in seq_along(seq_range)) {
    p <- seq_range[i]

    # Calculate nb using the provided formula
    nb <- TP / n - (FP) / n * (p / (100 - p))
    # Update n_values vector with nb
    n_values[i] <- nb
  }

  # Plot
  ggplot() +
    geom_line(mapping = aes(x = seq_range, 
                            y = n_values, 
                            col="index"), linewidth=1.2) +  # test
    geom_line(mapping = aes(x = c(0, input$prev), y = c(input$prev/100, 0), col="test_all"), linewidth=1.2) +  # treat all
    geom_line(mapping = aes(x = c(input$range[1], input$range[2]), y = 0, col="test_none"), linewidth=1.2) +  # treat none
    ylab("Net benefit\n") +
    xlab("\nThreshold probability (%)") +
    scale_x_continuous(breaks = seq(input$range[1], input$range[2], 5)) +
    coord_fixed(ratio = 100, ylim = c(0,0.5), xlim = c(input$range[1], input$range[2])) +
    ggtitle(paste("Clinical utility of", input$test, "in", input$cond)) +
    scale_colour_manual(name = 'Strategy', 
                      values =c('index'='#f1184c','test_all'='black', 'test_none'="#f7dc66"), 
                      labels = c(input$test, paste(input$ref, "to all"), paste(input$ref, "to none"))) +
    theme_economist_white(gray_bg = F, base_size = 14, base_family = "Atkinson Hyperlegible")
})


```
